---
title: "p8105_hw3_kr3145"
author: "Kallan Roan"
date: "2025-10-03"
output: github_document
---

Load libraries

```{r}
library(tidyverse)
library(lubridate)
library(patchwork)
library(p8105.datasets)
```

Preset settings for ggplot

```{r}
knitr::opts_chunk$set(
  fig.width = 12,
  fig.height = 12,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1

Load Instacart Data

```{r}
data("instacart")
```

`instacart` dataset has `r nrow(instacart)` observations, each row representing a product from an order. There are `r instacart |> summarize(n_distinct(order_id))` unique orders in the dataset. There are `r ncol(instacart)` variables. `eval_set` represents the which evaluation set this orders belongs in, which this data specifically only includes data from the "`r instacart |> pull(eval_set) |> unique()`" eval_set. 

Some key variables include `order_dow` and `order_hour_of_day`, representing the day of the week and hour of the day the order was placed respectively. `days_since_prior_order` order represents the number of days since the last order, capped at `r instacart |> pull(days_since_prior_order) |> max()` and NA if this is the customer's first order. `product_name`, `aisle_name`, and `department` represents the product, aisle, and department name respectively. `aisle_id`, `department_id`, `order_id`, `product_id`, and `user_id` represents the aisle, department, product, and user id respectively. `add_to_card_order` represents the order in which each product was added to cart. `reordered` indicates 1 if the product has been ordered in the past and 0 otherwise. The average order day is `r instacart |> pull(order_dow) |> mean() |> round()`, representing Thursday. The average order time is 
`r instacart |> pull(order_hour_of_day) |> mean() |> round()`:00. For the customers who have ordered from instacart previously, the average days since the last order is 
`r instacart |> pull(days_since_prior_order) |> mean() |> round()` days.

##### How many aisles are there, and which aisles are the most items ordered from?

Calculate which aisles have the most items ordered

```{r}
num_aisle_insta = 
  instacart |> 
  group_by(aisle) |> 
  summarize(
    n = n() # counts observations for each group
  ) |> 
  arrange(desc(n)) #arrange in desc order
```

There are `r nrow(num_aisle_insta)` different aisles. ``r num_aisle_insta |> slice(1) |> pull(aisle)`` aisle has the most items ordered of `r num_aisle_insta |> slice(1) |> pull(n)` items.

##### Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

Create plot

```{r}
num_aisle_insta |> 
  filter(n > 10000) |> #keep those with 10000 orders
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_col() +
  labs(
    x = "Aisle",
    y = "Number of Items Ordered", 
    title = "Number of Items Ordered in Each Aisle \nfor >10,000 items ordered", 
  ) +
  coord_flip()
 # theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

##### Make a table showing the three most popular items in each of the aisles “baking ingredients”, 
“dog food care”, and “packaged vegetables fruits”. Include the number of times each item is
ordered in your table.

```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |> 
  group_by(aisle, product_name) |> 
  summarize(
    items_ordered_count = n()
  ) |> 
  arrange(aisle, desc(items_ordered_count)) |> 
  slice_head(n = 3) |> 
  knitr::kable()
```

##### Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are 
ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r}
instacart |> 
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |> 
  group_by(product_name, order_dow) |> 
  summarize(
    mean_hour_of_day = round(mean(order_hour_of_day), 2)
  ) |> 
  pivot_wider(
    names_from = order_dow,
    values_from = mean_hour_of_day
  ) |> 
  rename( 
    sunday = `0`,
    monday = `1`,
    tuesday = `2`, 
    wednesday = `3`, 
    thursday = `4`, 
    friday = `5`, 
    saturday = `6`
  ) |> 
  knitr::kable()
```

# Problem 2

Read and clean Zip Codes.csv

```{r}
zip_code_df = 
  read_csv("data/Zip Codes.csv") |> 
  janitor::clean_names() |> 
  mutate(
    borough = case_match(
      county,
      "Bronx" ~ "Bronx",
      "Kings" ~ "Brooklyn",
      "New York" ~ "Manhattan",
      "Queens" ~ "Queens",
      "Richmond" ~ "Staten Island"
    )
  ) |> 
  select(zip_code, borough, neighborhood) |> 
  distinct(zip_code, .keep_all = TRUE)
```

Read and clean Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv

```{r}
zip_zori_df =
  read_csv("data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv") |> 
  janitor::clean_names() |> 
  pivot_longer(
    col = x2015_01_31:x2024_08_31,
    names_to = "month",
    values_to = "price",
    names_prefix = "x"
  ) |> 
  rename(zip_code = region_name) |> 
  mutate(
    month = as_date(month),
    year = floor_date(month, unit = "year"),
    zip_code = as.numeric(zip_code)) |> 
  select(-county_name)
```

Merge zip_zori_df and zip_code_df

```{r}
combined_zip_zori_df = 
  left_join(zip_zori_df, zip_code_df, by = "zip_code")
```

##### There are 116 months between January 2015 and August 2024. How many ZIP codes are observed 116 times? How many are observed fewer than 10 times? Why are some ZIP codes are observed rarely and others observed in each month?

Calculate number of zip codes observed 116 times

```{r}
combined_zip_zori_df |> 
  drop_na(price) |> 
  group_by(zip_code) |> 
  summarize(
    observed_per_zip_code = n()
  ) |> 
  count(observed_per_zip_code) |> 
  arrange(desc(observed_per_zip_code)) |> 
  filter(observed_per_zip_code == 116)
```

48 zip codes are observed 116 times. 

Calculate number of zip codes observed fewer than 10 times

```{r}
#output only those observed less than 10 times
combined_zip_zori_df |> 
  drop_na(price) |> 
  group_by(zip_code) |> 
  summarize(
    observed_per_zip_code = n()
  ) |> 
  count(observed_per_zip_code) |> 
  filter(observed_per_zip_code < 10) |> 
  summarize(less_10_zips = sum(n))
```

There are 26 zip codes that are observed less than 10 times.

???
Some zip codes are observed rarely because there were no rentals in that area previously, and there have been  been more rentals available in that area in recent areas. At the same time, it may be possible that there were rentals in that area previously, but there were less to no rentals in recent years. Some rentals may be more available in certain times of the year, so there will be more or less data during those time periods. 
Some zip codes are observed every month because there are enough rentals readily available that a ZORI score is accurately calculated for each month. This is the case especially for boroughs and neighborhoods with a larger population, like Manhattan.

##### Create a reader-friendly table showing the average rental price in each borough and year (not month). Comment on trends in this table.

```{r}
combined_zip_zori_df |> 
  group_by(borough, year) |> 
  summarize(
    mean = mean(price, na.rm = TRUE)
  ) |> 
  mutate(year = year(year)) |> 
  pivot_wider(
    names_from = year,
    values_from = mean
  ) |> 
  knitr::kable()
```

On average, the average yearly rental price has increased overtime for each borough. The average yearly rental price for Manhattan has increased the most, with the average yearly rental increased by over $1,000 from 2015 to 2024. Except for Staten Island, there is data for average yearly rental for all years between 2015 to 2024. Staten Island has no data on rental price from 2015 to 2019. 

##### Make a plot showing NYC Rental Prices within ZIP codes for all available years. Your plot should facilitate comparisons across boroughs. Comment on any significant elements of this plot.

```{r}
ggp_zip_price_all =
combined_zip_zori_df |> 
  drop_na(price) |> 
  group_by(borough, year) |> 
  summarize(
    mean_price = mean(price, na.rm = TRUE)
  ) |> 
  ggplot(aes(x = year, y = mean_price, color = borough)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(
    x = "Year",
    y = "Mean NYC Rental Prices", 
    title = "Mean NYC Rental Prices for all Available Years By Borough", 
    caption = "Data from Zillow"
  ) 

ggp_zip_price_all
```

On average, NYC rental prices has increased over the years. The plot shows that the change in price increases significantly starting 2021. The order in average rental prices across boroughs has stayed consistent throughout the years, with Bronx having the lowest average rental prices, followed by Staten Island, Queens, Brooklyn, and finally  Manhattan having the highest average rental prices in NYC. The plot also shows that there is only data for rental prices in Staten Island starting 2020. 

##### Compute the average rental price within each ZIP code over each month in 2023. Make a reader-friendly plot showing the distribution of ZIP-code-level rental prices across boroughs; put differently, your plot should facilitate the comparison of the distribution of average rental prices across boroughs. Comment on this plot.

Create plot to show the distribution of ZIP-code-level rental prices across boroughs

```{r}
ggp_zip_price_2023 = 
combined_zip_zori_df |> 
  filter(year(year) == 2023) |> 
  #group_by(borough, zip_code, month) |> 
  mutate(borough = forcats::fct_reorder(borough, price)) |> 
  ggplot(aes(x = borough, y = price)) +
  geom_boxplot() +
  labs(
    x = "Borough",
    y = "Mean NYC Rental Prices", 
    title = "Mean NYC Rental Prices in 2023 By Borough", 
    caption = "Data from Zillow"
  ) 

ggp_zip_price_2023
```

The boxplot shows that the order of boroughs from lowest to highest average rental prices remained the same, even when only looking at data from 2023. Staten island has the lowest range of rental prices, while Manhattan has the largest range of rental prices, having a rental prices as low as over \$2,000 to as high as over \$8,000. 

##### Combine the two previous plots into a single graphic, and export this to a results folder in your repository.

```{r}
combine_ggp_zip =
  ggp_zip_price_all / ggp_zip_price_2023 

ggsave("results/combine_ggp_zip.png", combine_ggp_zip, bg = "white")
```

The combined plot is located in a folder titled "results."

# Problem 3

Load and clean nhanes_covar.csv

```{r}
nhanes_covar_df =
  read_csv("data/nhanes_covar.csv",
           skip = 4) |> 
  janitor::clean_names() |> 
  drop_na() |> # drop those with missing demographic data 
  filter(age >= 21) # keep participants >= 21 years old; exclude participants less than 21 years of age
```

Load and clean nhanes_accel.csv

```{r}
nhanes_accel_df =
  read_csv("data/nhanes_accel.csv") |> 
  janitor::clean_names() 
```

Merge nhanes_covar.csv and nhanes_accel.csv

```{r}
nhanes_covar_accel_df =
  left_join(nhanes_covar_df, nhanes_accel_df, by = "seqn") |> 
  janitor::clean_names() |> 
  pivot_longer(
    cols = min1:min1440,
    names_to = "minute",
    values_to = "mims",
    names_prefix = "min"
  ) |> 
  mutate(
    sex = factor(case_match( # convert sex from numeric to factor
      sex,
      1 ~ "male", # add labels
      2 ~ "female"), 
      levels = c("male", "female")),
    education = factor(case_match( # convert education from numeric to factor
      education,
      1 ~ "less than high school", # add labels
      2 ~ "high school equivalent",
      3 ~ "more than high school"),
      levels = c("less than high school", "high school equivalent", "more than high school")),
    minute = as.numeric(minute) # convert minute from character to numeric
  )
```

##### Produce a reader-friendly table for the number of men and women in each education category 

```{r}
nhanes_covar_accel_df |> 
  group_by(sex, education) |> 
  summarize(
    n = n_distinct(seqn)
  ) |> 
  pivot_wider(
    names_from = education,
    values_from = n
  ) |> 
  knitr::kable()
```

There are more individuals who are in the more than high school education category compared to other education categories, even when stratified by male and female. Among males, there is a larger number of individuals who have a high school equivalent compared to less than high school. On the other hand, among females, there is a higher number of individuals who are categorized in less than high school compared to high school equivalent education category.

##### Create a visualization of the age distributions for men and women in each education category.

```{r}
nhanes_covar_accel_df |>
  distinct(seqn, sex, education, age) |>  
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 5) +
  facet_grid(sex ~ education) +
  labs(
    x = "Age (years)",
    y = "Count", 
    title = "Distribution of Age by Sex and Education", 
    caption = "Data from NHANES"
  ) 
```

The histogram shows that among males, there is a relatively even distribution of age for each education category. Among females, there is also a relatively even distribution of age for those in the less than high school and high school equivalent education categories. For females in the more than high school education category, the distribution appears slightly right skewed, with more females below 40's. 

##### Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

```{r}
nhanes_covar_accel_df |>
  group_by(seqn, sex, education, age) |> 
  summarise(
    total_activity = sum(mims) # calculate total activity by seqn
  ) |> 
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ education) +
  labs(
    x = "Age (years)",
    y = "Total Activity (minutes)", 
    title = "Distribution of Total Activity by Sex and Education", 
    caption = "Data from NHANES"
  ) 
```

Across all sexes and education categories, as age increases, the total activity in minutes decreases, on average. For those in the high school equivalent and more than high school education categories, females have higher total activity over the day compared to males. In the less than high school education category, females have higher total activity recorded compared to males in those 40 years old and younger. However, for those older than 40 years in the less than high school category, males recorded higher total activity compared females. 

##### Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.

```{r}
nhanes_covar_accel_df |>
  group_by(minute, sex, education) |> 
  summarize(
    activity_time = mean(mims)
  ) |> 
  ggplot(aes(x = minute, y = activity_time, color = sex)) +
  geom_point(alpha = 0.2) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ education) +
  labs(
    x = "Time in 24-hour Day (min)",
    y = "Activity time (min)", 
    title = "Activity Time in 24-hour Day by Sex and Education", 
    caption = "Data from NHANES"
  ) 
```

Across all sexes and education categories, the activity time in minutes is the lowest during the earliest and latest time of the 24-hour day, with more activity time during the day between 500 and 1000 minutes. There is a decrease in activity between 0 to 250 minutes, followed by a sharp increase in activity time until 500 minutes, and a sharp decrease after 1000 minutes across all sexes and education categories. On average, females have higher activity time throughout the day compared to males across all education categories. It also appears that as an individual has a higher education, there is a larger increase in gap of activity time between males and females. More specifically, while the difference in activity times between males and females does not appear as significant among those in the less than high school education category, those in more than high school education category have a larger difference in activity time between females and males. The plots also show that those with high school equivalent education have the lowest average activity time during the day compared to those with less than high school and more than high school education. 