---
title: "p8105_hw3_kr3145"
author: "Kallan Roan"
date: "2025-10-03"
output: github_document
---

Libraries
```{r}
library(tidyverse)
library(p8105.datasets)
```

Preset settings for ggplot

```{r}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


# Problem 1

Load Instacart Data
```{r}
data("instacart")
```

# Short description of dataset
- size and structure of data
- describe key variables
- give illustrative examples of observations

`instacart` dataset has `r nrow(instacart)` observations, each row representing a product of from an order. 
`r ncol(instacart)` variables.  

How many aisles are there, and which aisles are the most items ordered from?

```{r}
num_aisle_insta = 
  instacart |> 
  group_by(aisle) |> 
  summarize(
    n = n() # counts observations for each group
  ) |> 
  arrange(desc(n)) #arrange in desc order
```

There are `r nrow(num_aisle_insta)` different aisles. ``r pull(slice(num_aisle_insta, 1), aisle)`` aisle has the most items ordered of `r max(pull(num_aisle_insta, n))`.

Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

????Arrange aisle sensibly??

```{r}
#plot_num_aisle_insta =
  num_aisle_insta |> 
  filter(n > 10000) |> #keep those with 10000 orders
  ggplot(aes(x = aisle, y = n)) + 
  geom_col() +
  coord_flip() +
  labs(
    x = "Asile",
    y = "Number of Items Ordered", 
    title = "Number of Items Ordered in Each Aisle", 
  ) 
```

Make a table showing the three most popular items in each of the aisles “baking ingredients”, 
“dog food care”, and “packaged vegetables fruits”. Include the number of times each item is
ordered in your table.

```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |> 
  group_by(aisle, product_name) |> 
  summarize(
    n = n()
  ) |> 
  arrange(aisle, desc(n)) |> 
  slice_head(n = 3) |> 
  knitr::kable()
```

Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are 
ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r}

instacart |> 
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |> 
  group_by(product_name, order_dow) |> 
  summarize(
    mean_hour_of_day = mean(order_hour_of_day)
  ) |> 
  pivot_wider(
    names_from = order_dow,
    values_from = mean_hour_of_day
  ) |> 
  rename( 
    sunday = `0`,
    monday = `1`,
    tuesday = `2`, 
    wednesday = `3`, 
    thursday = `4`, 
    friday = `5`, 
    saturday = `6` ) |> 
  knitr::kable()
```

# Problem 2

Read and clean Zip Codes.csv

```{r}
zip_code_df =
  read_csv("data/Zip Codes.csv") |> 
  janitor::clean_names() |> 
  select(zip_code, neighborhood) |> 
  distinct(zip_code, .keep_all = TRUE)

```

Read and clean Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv

```{r}
zip_zori_df =
  read_csv("data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv") |> 
  janitor::clean_names() |> 
  pivot_longer(
    col = x2015_01_31:x2024_08_31,
    names_to = "date",
    values_to = "zori",
    names_prefix = "x"
  ) |> 
  separate(
    date, into = c("year", "month", "day"), sep = "_", convert = TRUE
  ) |> 
  rename(zip_code = region_name, county = county_name) |> 
  select(zip_code, county, year, month, zori) |> 
  mutate(
    county = str_trim(str_remove(county, "County")),
    month = month.name[month]
  ) 
```

Merge zip_zori_df and zip_code_df

```{r}
combined_zip_zori_df =
  left_join(zip_code_df, zip_zori_df, by = c("zip_code")) |> 
  janitor::clean_names()
```

There are 116 months between January 2015 and August 2024. How many ZIP codes are observed 116 times? How many are observed fewer than 10 times? Why are some ZIP codes are observed rarely 
and others observed in each month?

```{r}
combined_zip_zori_df |> 
  group_by(zip_code) |> 
  summarize(
    n = n()
  )
```

Create a reader-friendly table showing the average rental price in each borough and year (not month). Comment on trends in this table.

```{r}
combined_zip_zori_df |> 
  group_by(county, year) |> 
  summarize(
    mean = mean(zori, na.rm = TRUE)
  ) |> 
  pivot_wider(
    names_from = year,
    values_from = mean
  ) |> 
  knitr::kable()
```

Make a plot showing NYC Rental Prices within ZIP codes for all available years. Your plot should facilitate comparisons across boroughs. Comment on any significant elements of this plot.

Compute the average rental price within each ZIP code over each month in 2023. Make a reader-friendly plot showing the distribution of ZIP-code-level rental prices across boroughs; put differently, your plot should facilitate the comparison of the distribution of average rental prices across boroughs. Comment on this plot.

Combine the two previous plots into a single graphic, and export this to a results folder in your repository.



# Problem 3

Load and clean nhanes_covar.csv

```{r}
nhanes_covar_df =
  read_csv("data/nhanes_covar.csv",
           skip = 4) |> 
  janitor::clean_names() |> 
  drop_na() |> # drop those with missing demographic data 
  filter(age >= 21) # keep participants >= 21 years old; exclude participants less than 21 years of age
```

Load and clean nhanes_accel.csv

```{r}
nhanes_accel_df =
  read_csv("data/nhanes_accel.csv") |> 
  janitor::clean_names() 
```

Merge nhanes_covar.csv and nhanes_accel.csv

```{r}
nhanes_covar_accel_df =
  left_join(nhanes_covar_df, nhanes_accel_df, by = c("seqn")) |> 
  janitor::clean_names() |> 
  pivot_longer(
    cols = min1:min1440,
    names_to = "minute",
    values_to = "mims",
    names_prefix = "min"
  ) |> 
  mutate(
    sex = factor(case_match( # convert sex from numeric to factor
      sex,
      1 ~ "male", # add labels
      2 ~ "female"), 
      levels = c("male", "female")),
    education = factor(case_match( # convert education from numeric to factor
      education,
      1 ~ "less than high school", # add labels
      2 ~ "high school equivalent",
      3 ~ "more than high school"),
      levels = c("less than high school", "high school equivalent", "more than high school")),
    minute = as.numeric(minute) # convert minute from character to numeric
  )
```

notes: delete later
- seqn: numeric 
- sex: numeric -> factor
- age: numeric
- bmi: numeric
- edu: numeric -> factor
- minute: character -> numeric
- mims: numeric

Produce a reader-friendly table for the number of men and women in each education category 

```{r}
nhanes_covar_accel_df |> 
  group_by(sex, education) |> 
  summarize(
    n = n_distinct(seqn)
  ) |> 
  pivot_wider(
    names_from = education,
    values_from = n
  ) |> 
  knitr::kable()


```

??????Comment on these items.

Create a visualization of the age distributions for men and women in each education category.

```{r}
nhanes_covar_accel_df |>
  distinct(seqn, sex, education, age) |>  
  ggplot(aes(x = age)) +
  geom_histogram() +
  facet_grid(sex ~ education)
```

??????Comment on these items.



Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

```{r}
nhanes_covar_accel_df |>
  group_by(seqn, sex, education, age) |> 
  summarise(
    total_activity = sum(mims) # calculate total activity by seqn
  ) |> 
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ education)
```


Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.

```{r}
nhanes_covar_accel_df |>
  group_by(minute, sex, education) |> 
  summarize(
    activity_time = mean(mims)
  ) |> 
  ggplot(aes(x = minute, y = activity_time, color = sex)) +
  geom_point(alpha = 0.1) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ education)
```

