---
title: "p8105_hw3_kr3145"
author: "Kallan Roan"
date: "2025-10-03"
output: github_document
---

Libraries
```{r}
library(tidyverse)
library(p8105.datasets)
```

# Problem 1

Load Instacart Data
```{r}
data("instacart")
```

# Short description of dataset
- size and structure of data
- describe key variables
- give illustrative examples of observations

`instacart` dataset has `r nrow(instacart)` observations, each row representing a product of from an order. 
`r ncol(instacart)` variables.  

How many aisles are there, and which aisles are the most items ordered from?

```{r}
num_aisle_insta = 
  instacart |> 
  group_by(aisle) |> 
  summarize(
    n = n() # counts observations for each group
  ) |> 
  arrange(desc(n)) #arrange in desc order
```

There are `r nrow(num_aisle_insta)` different aisles. ``r pull(slice(num_aisle_insta, 1), aisle)`` aisle has the most items ordered of `r max(pull(num_aisle_insta, n))`.

Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

????Arrange aisle sensibly??

```{r}
#plot_num_aisle_insta =
  num_aisle_insta |> 
  filter(n > 10000) |> #keep those with 10000 orders
  ggplot(aes(x = aisle, y = n)) + 
  geom_col() +
  coord_flip() +
  labs(
    x = "Asile",
    y = "Number of Items Ordered", 
    title = "Number of Items Ordered in Each Aisle", 
  ) 
```

Make a table showing the three most popular items in each of the aisles “baking ingredients”, 
“dog food care”, and “packaged vegetables fruits”. Include the number of times each item is
ordered in your table.

```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |> 
  group_by(aisle, product_name) |> 
  summarize(
    n = n()
  ) |> 
  arrange(aisle, desc(n)) |> 
  slice_head(n = 3) |> 
  knitr::kable()
```

Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are 
ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r}

instacart |> 
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |> 
  group_by(product_name, order_dow) |> 
  summarize(
    mean_hour_of_day = mean(order_hour_of_day)
  ) |> 
  pivot_wider(
    names_from = order_dow,
    values_from = mean_hour_of_day
  ) |> 
  rename( 
    sunday = `0`,
    monday = `1`,
    tuesday = `2`, 
    wednesday = `3`, 
    thursday = `4`, 
    friday = `5`, 
    saturday = `6` ) |> 
  knitr::kable()
```

